\section{Conclusion}
\label{sec:conclusion}

We presented BDI-LLM, a framework that integrates the Belief-Desire-Intention cognitive architecture with large language models and multi-layer formal verification for automated planning. Our approach addresses the well-documented unreliability of LLMs in multi-step planning tasks through three key contributions: (1) domain-aware structured prompting that converts PDDL problems into rich natural language representations with explicit constraints and worked examples, (2) a BDI-structured output format that constrains LLM generation to well-formed directed acyclic graphs, and (3) a multi-layer verification pipeline incorporating structural checks, VAL-based symbolic verification, and an error-driven repair loop.

Evaluated on 1,270 PlanBench instances across Blocksworld, Logistics, and Depots, our framework achieves 99.6\% overall accuracy---100\% on Blocksworld, 99.6\% on Logistics, and 99.4\% on Depots. These results represent improvements of 65--95 percentage points over prior LLM planning baselines on the same benchmark. The VAL repair loop proves essential, successfully correcting 107 out of 111 initially invalid plans (96.4\% repair success rate).

\subsection{Future Work}

Several directions merit further investigation:

\begin{itemize}
    \item \textbf{Automated Domain Adaptation}: Developing methods to automatically generate domain-specific prompts from PDDL domain specifications, reducing the manual engineering required for new domains.

    \item \textbf{Scaling to Larger Instances}: Evaluating performance on problems with 100+ objects and investigating hierarchical decomposition strategies for complex instances that exceed LLM context limits.

    \item \textbf{Multi-Model Evaluation}: Benchmarking the framework across diverse LLMs (GPT-4o, Claude Opus, Llama 3) to assess generalizability and identify model-specific optimization opportunities.

    \item \textbf{Advanced Repair Strategies}: Exploring plan decomposition, subgoal-based regeneration, and ensemble repair methods to address the remaining 0.4\% failure cases.

    \item \textbf{Real-World Deployment}: Extending the framework to robotics task planning and software agent orchestration, where formal verification of LLM-generated plans is critical for safety.
\end{itemize}
